---
title: "6.2 R Code"
output: 
  pdf_document:
    latex_engine: xelatex
---

# Install Packages
```{r, inclued=FALSE}
# install.packages(c("dplyr", "tidyr", "ggplot2", ""plotly","Hmisc", "corrplot","dplyr", "corrplot", "factoextra", "broom", "ggrepel", "purrr", "psych"))

library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(corrplot)
library(factoextra)
library(broom)
library(ggrepel)
library(purrr)
library(psych)
```

# 1. Prepare Data
## 1.1 Load ABIDE Dataset
```{r}
dt <- load("/Users/macbook/Desktop/ABIDE_YALE.RData")
```

## 1.2 Check Data
```{r}
print(dt)
```
"dt": include 2 datasets, "YALE_fmri" & "YALE_demo_var"

```{r}
summary(YALE_fmri)
str(YALE_fmri[1])
```
"YALE_fmri": list of 47 matrices, each 196x110

```{r}
missing_fmri <- sapply(YALE_fmri, function(mat) sum(is.na(mat)))

cat("\n=== fMRI Data Missing Value Summary ===\n")
cat("Total missing values:", sum(missing_fmri), "\n")
cat("Number of subjects with missing values:", sum(missing_fmri > 0), "\n")
```

```{r}
str(YALE_demo_var)
```
"YALE_demo_var": data frame with 47 rows, 3 variables

```{r}
missing_demo <- colSums(is.na(YALE_demo_var))

cat("Census Table Missing Value Summary:\n")
print(missing_demo)
```
There is no missing data

## 1.3 Time series preprocessing
```{r}
YALE_fmri_processed <- lapply(1:length(YALE_fmri), function(i) {
  subject_mat <- YALE_fmri[[i]]
  subject_mat <- as.matrix(subject_mat)
  storage.mode(subject_mat) <- "double"
  
  global_mean <- mean(subject_mat)
  subject_mat <- subject_mat - global_mean
  
  min_val <- min(subject_mat)
  subject_mat_shifted <- subject_mat - min_val + 1e-3
  log_mat <- log(subject_mat_shifted)
  
  processed_mat <- apply(log_mat, 2, function(ts) {
    (ts - mean(ts)) / sd(ts)
  })
  
  if (i %in% sample(1:47, 3)) {
    sample_regions <- sample(1:110, 5)
    par(mfrow = c(2, 5), mar = c(3,3,2,1), oma = c(0,0,2,0))
    for (r in sample_regions) {
      hist(subject_mat[, r], breaks = 30, main = paste("Region", r, "-Original"), 
           xlab = "", col = "skyblue")
      hist(processed_mat[, r], breaks = 30, main = paste("Region", r, "-After"), 
           xlab = "", col = "salmon")
    }
    mtext(paste("Subject", i, "Pretreatment Quality Check "), outer = TRUE, cex = 1.2)
  }
  
  return(processed_mat)
})
```

# 2 Data Informations
```{r}
head(YALE_demo_var)
summary(YALE_demo_var)
```
Note: 
  DX_GROUP → Diagnosis (1 = Autism, 2 = Control).
  SEX → Gender (1 = Male, 2 = Female).

```{r}
demo_summary <- YALE_demo_var %>%
  group_by(DX_GROUP = factor(DX_GROUP, labels = c("Autism", "Control"))) %>%
  summarise(
    N = n(),
    Age_Mean = mean(AGE_AT_SCAN, na.rm = TRUE),
    Age_SD = sd(AGE_AT_SCAN, na.rm = TRUE),
    Male = sum(SEX == 1),
    Female = sum(SEX == 2)
  )

print(demo_summary)
```

```{r}
subject_summaries <- lapply(YALE_fmri_processed, function(mat) {
  apply(mat, 2, function(ts) {
    c(
      Mean = mean(ts),
      SD = sd(ts),
      Q25 = quantile(ts, 0.25),
      Median = quantile(ts, 0.5),
      Q75 = quantile(ts, 0.75)
    )
  }) %>% t() %>% as.data.frame()
})

for (i in 1:length(subject_summaries)) {
  subject_summaries[[i]]$Subject_ID <- i
  subject_summaries[[i]]$Group <- YALE_demo_var$DX_GROUP[i]
}

summary_df <- do.call(rbind, subject_summaries)

summary_df$Group <- factor(summary_df$Group, levels = c(1,2), labels = c("Autism", "Control"))

group_summary <- summary_df %>%
  group_by(Group) %>%
  summarise(across(.cols = -Subject_ID, .fns = list(Mean = mean, SD = sd), .names = "{.col}_{.fn}"))

print(group_summary)
```

# 3 Data Analysis
## 3.1 Analyze by demographic variables
```{r}
ggplot(YALE_demo_var, aes(x = factor(DX_GROUP, labels = c("Autism", "Control")), 
                          y = AGE_AT_SCAN, 
                          color = factor(SEX))) + 
  geom_point() + 
  scale_color_manual(values = c("1" = "#4D97CD", "2" = "#DB6968"), 
                     labels = c("Male", "Female")) +
  labs(x = "Diagnosis", y = "Age at the time of the scan", color = "Sex") + 
  theme_minimal()
```
According to this graph, it can be preliminarily determined that Autism has no strong association with gender and age. Further verification follows:

```{r}
shapiro_autism <- shapiro.test(YALE_demo_var$AGE_AT_SCAN[YALE_demo_var$DX_GROUP == "1"])
shapiro_control <- shapiro.test(YALE_demo_var$AGE_AT_SCAN[YALE_demo_var$DX_GROUP == "2"])

cat("Autism group normality p-values:", shapiro_autism$p.value, "\n")
cat("Control group normality p-values:", shapiro_control$p.value, "\n")

# If both groups are normal, the T-test is used. Otherwise, the Mann-Whitney U test is used
if (shapiro_autism$p.value > 0.05 & shapiro_control$p.value > 0.05) {
  t_test <- t.test(AGE_AT_SCAN ~ DX_GROUP, data = YALE_demo_var)
  cat("T-test results:\n")
  print(t_test)
} else {
  wilcox_test <- wilcox.test(AGE_AT_SCAN ~ DX_GROUP, data = YALE_demo_var)
  cat("Mann-Whitney U test results:\n")
  print(wilcox_test)
}

ggplot(YALE_demo_var, 
       aes(
         x = factor(DX_GROUP, labels = c("Autism", "Control")),
         y = AGE_AT_SCAN, 
         fill = factor(DX_GROUP)
       )) +
  geom_boxplot() +
  scale_fill_manual(
    values = c("#FF6B6B", "#4ECDC4"),
    labels = c("Autism", "Control"),
    name = "Group"
  ) +
  labs(
    x = "Group", 
    y = "Age", 
    title = "Age Distribution of Autism and Control Group"
  ) +
  theme_minimal()

```
There was no significant difference in age.

```{r}
sex_table <- table(YALE_demo_var$SEX, YALE_demo_var$DX_GROUP)

# Chi Square test (if expected frequency >= 5)
if (all(chisq.test(sex_table)$expected >= 5)) {
  chisq_test <- chisq.test(sex_table)
  cat("Chi-square test results:\n")
  print(chisq_test)
} else {
  fisher_test <- fisher.test(sex_table)
  cat("Fisher test results:\n")
  print(fisher_test)
}

ggplot(YALE_demo_var, aes(x = factor(DX_GROUP, labels = c("Autism", "Control")), 
       fill = factor(SEX))) +
  geom_bar(position = "fill") +
  scale_fill_manual(
    values = c("#4D97CD", "#DB6968"),
    labels = c("Male", "Female"),
    name = "Sex"
  ) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Group", 
    y = "Sex Ratio", 
    title = "Autism and Control Group Sex Ratio"
  ) +
  theme_minimal()
```
There was no significant difference in sex distribution.

## 3.2 Consider Brain Activities
### 3.2.1 Overall Comparison
```{r}
autism_mats <- YALE_fmri[YALE_demo_var$DX_GROUP == 1]
autism_cor_mean <- apply(simplify2array(lapply(autism_mats, cor)), 1:2, mean)

control_mats <- YALE_fmri[YALE_demo_var$DX_GROUP == 2]
control_cor_mean <- apply(simplify2array(lapply(control_mats, cor)), 1:2, mean)

par(
  mfrow = c(1, 2), 
  mar = c(0, 0, 5, 0),
  oma = c(0, 0, 2, 0)
)

add_title <- function(text, cex = 0.8, line = 3.5) {
  title(
    main = text,
    cex.main = cex,
    line = line
  )
}

corrplot(
  autism_cor_mean, 
  method = "color", 
  tl.pos = "n"
)
add_title("Average Connection of Autism Group")

corrplot(
  control_cor_mean, 
  method = "color", 
  tl.pos = "n"
)
add_title("Average Connection of Control Group")

diff_matrix <- autism_cor_mean - control_cor_mean
corrplot(
  diff_matrix, 
  method = "color", 
  tl.pos = "n"
)
add_title("Connection Difference (Autism-Control)")
```
Brain activity was similar between the two groups, however, Autistic group had slightly higher average activity intensity.

### 3.2.2 Brain Region Compare
#### Compare by Mean
```{r}
get_region_mean <- function(mat) {
  data.frame(
    Region = 1:ncol(mat),
    Mean = colMeans(mat, na.rm = TRUE)
  )
}

mean_data <- lapply(YALE_fmri, get_region_mean) %>%
  bind_rows(.id = "SubjectID") %>%
  mutate(SubjectID = as.numeric(SubjectID))

demo_clean <- YALE_demo_var %>%
  mutate(
    SubjectID = row_number(),
    DX_GROUP = factor(DX_GROUP, labels = c("Autism", "Control"))
  ) %>%
  select(SubjectID, DX_GROUP)

mean_data <- mean_data %>%
  left_join(demo_clean, by = "SubjectID")

region_stats <- mean_data %>%
  group_by(Region, DX_GROUP) %>%
  summarise(
    Group_Mean = mean(Mean, na.rm = TRUE),
    Group_SD = sd(Mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = DX_GROUP,
    values_from = c(Group_Mean, Group_SD),
    names_glue = "{.value}_{DX_GROUP}"
  ) %>%
  mutate(
    Diff_Mean = Group_Mean_Autism - Group_Mean_Control
  )

all_pvalues <- sapply(1:110, function(r) {
  dat <- mean_data %>% filter(Region == r)
  if (nrow(dat) < 2) return(NA)
  t.test(Mean ~ DX_GROUP, data = dat)$p.value
})

region_stats <- region_stats %>%
  mutate(
    p.value = all_pvalues,
    p.adj = p.adjust(p.value, method = "fdr")
  )

region_stats <- region_stats %>%
  mutate(
    Significance = case_when(
      p.adj < 0.05 & abs(Diff_Mean) > 0.3 ~ "FDR < 0.05 & |Diff| > 0.3",
      p.adj < 0.05 ~ "FDR < 0.05",
      TRUE ~ "Non-significant"
    )
  )

ggplot(region_stats, aes(x = Diff_Mean, y = -log10(p.adj))) +
  geom_point(aes(color = Significance, size = abs(Diff_Mean)), alpha = 0.7) +
  scale_color_manual(
    values = c(
      "FDR < 0.05 & |Diff| > 0.3" = "red",
      "FDR < 0.05" = "orange",
      "Non-significant" = "grey80"
    )
  ) +
  geom_vline(xintercept = c(-0.3, 0.3), linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_text_repel(
    data = subset(region_stats, p.adj < 0.05 & abs(Diff_Mean) > 0.3),
    aes(label = Region),
    size = 3,
    max.overlaps = 20,
    box.padding = 0.5
  ) +
  labs(
    x = "Mean difference between groups (Autism - Control)",
    y = "-log10(FDR correction p-value)",
    title = "Differences in Mean Signal Strength across all Brain Regions ",
    caption = "Dashed line: | difference | > 0.3 and FDR < 0.05"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

#### Compare by Range
```{r}
get_region_range <- function(mat) {
  data.frame(
    Region = 1:ncol(mat),
    Min = apply(mat, 2, min, na.rm = TRUE),
    Max = apply(mat, 2, max, na.rm = TRUE)
  )
}

range_data <- lapply(YALE_fmri, get_region_range) %>%
  bind_rows(.id = "SubjectID") %>%
  mutate(SubjectID = as.numeric(SubjectID))

demo_clean <- YALE_demo_var %>%
  mutate(
    SubjectID = row_number(),
    DX_GROUP = factor(DX_GROUP, labels = c("Autism", "Control"))
  ) %>%
  select(SubjectID, DX_GROUP)

range_data <- range_data %>%
  left_join(demo_clean, by = "SubjectID")

range_data <- range_data %>%
  mutate(Range = Max - Min)

region_diff <- range_data %>%
  group_by(Region, DX_GROUP) %>%
  summarise(
    Mean_Range = mean(Range, na.rm = TRUE),
    SD_Range = sd(Range, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = DX_GROUP,
    values_from = c(Mean_Range, SD_Range),
    names_glue = "{.value}_{DX_GROUP}"
  ) %>%
  mutate(
    Diff_Mean = Mean_Range_Autism - Mean_Range_Control
  )

all_pvalues <- sapply(1:110, function(r) {
  dat <- range_data %>% filter(Region == r)
  if (nrow(dat) == 0) return(NA)
  wilcox.test(Range ~ DX_GROUP, data = dat)$p.value
})

region_diff <- region_diff %>%
  mutate(
    p.value = all_pvalues,
    p.adj = p.adjust(p.value, method = "fdr")
  )

region_diff <- region_diff %>%
  mutate(
    Significance = case_when(
      p.adj < 0.05 & abs(Diff_Mean) > 0.5 ~ "FDR < 0.05 & |Diff| > 0.5",
      p.adj < 0.05 ~ "FDR < 0.05",
      TRUE ~ "Non-significant"
    )
  )

ggplot(region_diff, aes(x = Diff_Mean, y = -log10(p.adj))) +
  geom_point(aes(color = Significance, size = abs(Diff_Mean)), alpha = 0.7) +
  scale_color_manual(
    values = c(
      "FDR < 0.05 & |Diff| > 0.5" = "red",
      "FDR < 0.05" = "orange",
      "Non-significant" = "grey80"
    )
  ) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_text_repel(
    data = subset(region_diff, p.adj < 0.05 & abs(Diff_Mean) > 0.5),
    aes(label = Region),
    size = 3,
    max.overlaps = 20,
    box.padding = 0.5
  ) +
  labs(
    x = "Difference Between Groups (Autism - Control)",
    y = "-log10(FDR correction p-value)",
    title = "Differences in Signal Intensity Range across all Brain Regions",
    caption = "Dashed line: | difference | > 0.5 and FDR < 0.05"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
range_data <- range_data %>%
  mutate(Range = Max - Min)

region_stats <- range_data %>%
  group_by(Region, DX_GROUP) %>%
  summarise(
    Mean_Range = mean(Range, na.rm = TRUE),
    SD_Range = sd(Range, na.rm = TRUE)
  ) %>%
  pivot_wider(
    names_from = DX_GROUP,
    values_from = c(Mean_Range, SD_Range)
  )

region_diff <- region_stats %>%
  mutate(
    Diff_Mean = Mean_Range_Autism - Mean_Range_Control,
    Diff_SD = SD_Range_Autism - SD_Range_Control
  ) %>%
  arrange(desc(abs(Diff_Mean)))

head(region_diff, )
```

```{r}
region31 <- range_data %>% filter(Region == 31)

ggplot(region31, aes(x = DX_GROUP, y = Max, fill = DX_GROUP)) +
  geom_boxplot(width = 0.4, alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = DX_GROUP), width = 0.1, size = 2, alpha = 0.7) + 
  geom_point(aes(y = Min), color = "grey40", size = 2, alpha = 0.5) +
  scale_fill_manual(values = c("Autism" = "#E69F00", "Control" = "#56B4E9")) +
  scale_color_manual(values = c("Autism" = "#E69F00", "Control" = "#56B4E9")) +
  labs(
    title = "Comparison of Signal Intensity Range in Brain Region with the Largest Between-Group Difference",
    x = "Diagnostic Group", 
    y = "Signal strength",
    caption = "Color dot: maximum, Gray dot: minimum"
  ) +
  theme_minimal()
```

```{r}
top_region <- region_diff$Region[1]

target_data <- range_data %>% filter(Region == top_region)

wilcox_result <- wilcox.test(Range ~ DX_GROUP, data = target_data)

cat("Statistical Test Results for the Brain Region with the Largest Between-Group Difference\n")
print(wilcox_result)
```

### 3.2.3 Brain Regions Connection Compare
```{r}
compute_fc <- function(processed_mat) {
  stopifnot(all(abs(colMeans(processed_mat)) < 1e-6))
  stopifnot(all(abs(apply(processed_mat, 2, sd) - 1) < 1e-3))
  
  cor_mat <- cor(processed_mat)
  z_mat <- 0.5 * log((1 + cor_mat) / (1 - cor_mat))
  diag(z_mat) <- 0
  return(z_mat)
}

fc_mats <- lapply(YALE_fmri_processed, compute_fc)

demo_clean <- YALE_demo_var %>%
  mutate(
    SubjectID = row_number(),
    DX_GROUP = factor(DX_GROUP, labels = c("Autism", "Control"))
  ) %>%
  select(SubjectID, DX_GROUP)

autism_mats <- fc_mats[demo_clean$DX_GROUP == "Autism"]
control_mats <- fc_mats[demo_clean$DX_GROUP == "Control"]

autism_mean <- apply(simplify2array(autism_mats), 1:2, mean)
control_mean <- apply(simplify2array(control_mats), 1:2, mean)

epsilon <- 1e-6
relative_diff <- (autism_mean - control_mean) / 
                 ((abs(autism_mean) + abs(control_mean))/2 + epsilon)

relative_diff[lower.tri(relative_diff, diag = TRUE)] <- NA

actual_min <- floor(min(relative_diff, na.rm = TRUE))
actual_max <- ceiling(max(relative_diff, na.rm = TRUE))
scale_factor <- max(abs(actual_min), abs(actual_max))
relative_diff_scaled <- relative_diff / scale_factor

col_palette <- colorRampPalette(c("blue", "white", "red"))(100)

corrplot(
  relative_diff_scaled,
  method = "color",
  col = col_palette,
  tl.pos = "n",
  col.lim = c(-1, 1),
  title = "Standardized Differences between Brain Regions Connection",
  mar = c(0, 0, 2, 0))
```

#### Sorted by uncorrected p-value and effect size
```{r}
compute_fc <- function(mat) {
  cor(mat)
}

fc_mats <- lapply(YALE_fmri, compute_fc)

demo_clean <- YALE_demo_var %>%
  mutate(
    SubjectID = row_number(),
    DX_GROUP = factor(DX_GROUP, labels = c("Autism", "Control"))
  ) %>%
  select(SubjectID, DX_GROUP)

extract_connections <- function(mat) {
  data.frame(
    from = rep(1:110, each = 110),
    to = rep(1:110, times = 110),
    strength = as.vector(mat)
  ) %>%
    filter(from < to)
}

all_connections <- lapply(fc_mats, extract_connections) %>%
  bind_rows(.id = "SubjectID") %>%
  mutate(SubjectID = as.numeric(SubjectID)) %>%
  left_join(demo_clean, by = "SubjectID")

connection_stats <- all_connections %>%
  group_by(from, to, DX_GROUP) %>%
  summarise(
    mean_strength = mean(strength, na.rm = TRUE),
    sd_strength = sd(strength, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = DX_GROUP,
    values_from = c(mean_strength, sd_strength),
    names_glue = "{.value}_{DX_GROUP}"
  ) %>%
  mutate(
    diff_strength = mean_strength_Autism - mean_strength_Control
  )

p_values <- sapply(1:nrow(connection_stats), function(i) {
  conn <- connection_stats[i, ]
  autism_vals <- all_connections %>%
    filter(from == conn$from, to == conn$to, DX_GROUP == "Autism") %>%
    pull(strength)
  control_vals <- all_connections %>%
    filter(from == conn$from, to == conn$to, DX_GROUP == "Control") %>%
    pull(strength)
  if (length(autism_vals) < 2 || length(control_vals) < 2) return(NA)
  t.test(autism_vals, control_vals)$p.value
})

connection_stats <- connection_stats %>%
  mutate(
    p.value = p_values,
    p.adj = p.adjust(p.value, method = "fdr")
  )

connection_stats <- connection_stats %>%
  mutate(
    pooled_sd = sqrt(
      (sd_strength_Autism^2 * (sum(demo_clean$DX_GROUP == "Autism") - 1) +
       sd_strength_Control^2 * (sum(demo_clean$DX_GROUP == "Control") - 1)) /
      (sum(demo_clean$DX_GROUP == "Autism") + sum(demo_clean$DX_GROUP == "Control") - 2)
    ),
    cohen_d = diff_strength / pooled_sd
  )

top_connections <- connection_stats %>%
  arrange(p.value, desc(abs(cohen_d))) %>%
  slice_head(n = 10)

print(top_connections %>% select(from, to, diff_strength, cohen_d, p.value))
```

#### Sorted by effect size
```{r}
top_connections <- connection_stats %>%
  arrange(desc(abs(cohen_d))) %>%
  slice_head(n = 10)

print(top_connections %>% select(from, to, cohen_d, p.value, p.adj))
```

#### Sorted by p-value
```{r}
p_values <- sapply(1:nrow(connection_stats), function(i) {
  conn <- connection_stats[i, ]
  autism_vals <- all_connections %>%
    filter(from == conn$from, to == conn$to, DX_GROUP == "Autism") %>%
    pull(strength)
  control_vals <- all_connections %>%
    filter(from == conn$from, to == conn$to, DX_GROUP == "Control") %>%
    pull(strength)
  if (length(autism_vals) < 2 || length(control_vals) < 2) return(NA)
  wilcox.test(autism_vals, control_vals)$p.value
})

connection_stats <- connection_stats %>%
  mutate(
    p.value = p_values,
    p.adj = p.adjust(p.value, method = "fdr")
  )

top_connections <- connection_stats %>%
  filter(p.adj < 0.1) %>%
  arrange(p.adj, desc(abs(cohen_d))) %>%
  slice_head(n = 10)

top_effects <- connection_stats %>%
  arrange(desc(abs(cohen_d))) %>%
  slice_head(n = 10)

ggplot(top_effects, aes(x = reorder(paste(from, "-", to), cohen_d), 
                        y = cohen_d, fill = cohen_d > 0)) +
  geom_col() +
  scale_fill_manual(values = c("blue", "red")) +
  labs(
    x = "Brain Area Connection", 
    y = "Cohen's D Effect Size",
    title = "10 Brain Region Connections with the Largest Effect Size"
  ) +
  coord_flip() +
  theme_minimal()
```

# 4 Verification
```{r}
all_z <- unlist(lapply(fc_mats, function(mat) mat[upper.tri(mat)]))

ggplot(data.frame(Z = all_z), aes(x = Z)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "skyblue") +
  stat_function(fun = dnorm, args = list(mean = mean(all_z), sd = sd(all_z)), 
                color = "red", size = 1) +
  labs(title = "Functional Connection Fisher Z value distribution", 
       subtitle = "Red curve is normal distribution reference line") +
  theme_minimal()
```